Cryptanalysis of \MORUS

### 2018-07-12 AC18_rebuttal.txt

************************************************************************
* REBUTTAL *************************************************************
************************************************************************

We thank all reviewers for their positive comments.
We will take all editorial comments into account to prepare an updated version.
Below, we discuss some specific comments raised by the reviewers:

> A small drawback that I would add is that this paper would benefit from a
> concrete comparison with similar attacks, as the previous one that was found
> on AEGIS. While it does not low down the impact of the attack on a new scheme
> it could help to understand the differences for further cryptanalysis.

Broadly speaking, the attack on MORUS is similar to traditional linear attacks
on stream ciphers. It is closest to the attack on AEGIS. In both cases, the
approach is to build a linear trail linking ciphertext bits, while canceling the
contribution of inner state bits. How the trail is built depends on the state
update function, and differs significantly between AEGIS and MORUS, as the AEGIS
state update function relies on the parallel application of AES rounds, rather
than shift/AND/XOR operations.

### 2018-07-12 AC18_rebuttal.txt

https://secure.iacr.org/websubrev/ac2018/submit/rebuttal.php

Remember that the deadline for the rebuttals is July 13th, 11:45 UTC.

************************************************************************
* REBUTTAL *************************************************************
************************************************************************

We thank all reviewers for their positive comments.
We will take all editorial comments into account to prepare an updated version.
Below, we discuss some specific comments raised by the reviewers. We agree with
those comments and will update the paper with the relevant clarifications.

@@

> Moreover, the provided attack works either if the key is fixed or if the same
> plaintext is repeated a large number of times. While the former case is not a
> big limitation in practice even considering design recommendations, the later
> seems to be less probable in practice. Maybe examples of concrete scenarii
> should be emphasized to strengthen the proposal.

As discussed in Section 6, repeated plaintexts actually appear in many practical
applications due to fixed formats (headers, metadata, file formats) and often
include some repeated secret parts (password, authentication cookie, ...).
This scenario has been exploited in several practical attacks, including BEAST
(TLS), RC4 attacks (TLS, WPA), and Sweet32 (TLS, OpenVPN).

> In the plaintext recovery attack in the broad setting, it requires more
> samples than distinguishing attacks. So, the attack does not match the
> practical attack scenario such as BEAST.

Clearly, the complexity of the attack we propose is not practical, unlike the
attacks on legacy cryptosystems in other BEAST-style attacks.  However, the
attack complexity is less than suggested by the security claim of MORUS1280-256.
We are not aware of a distinguishing attack on MORUS with lower complexity than
our plaintext recovery attack. Note that MORUS is not a block cipher, and
distinguishing attacks on modes of operation (such as the CBC mode targeted in
the original BEAST attack) do not apply.

### 2018-09-06 AC2018_camera_ready/morusAC_00_Abstract.tex
Camera-ready version

  As our main result, we present a
  linear correlation in the keystream of full \MORUS, which can be used to distinguish its output
  from random and to recover some plaintext bits in the
  broadcast setting.  For \MORUS[1280], the correlation is $2^{-76}$,
  which can be exploited after around $2^{152}$ encryptions, less than
  would be expected for a 256-bit secure cipher.  For
  \MORUS[640], the same attack results in a correlation of $2^{-73}$, which does not
  violate the security claims of the cipher.

% More precisely, we show that a linear correlation exists between some bits of the plaintext and ciphertext. Moreover the bias depends purely on the plaintext, and not on the secret key nor the nonce, hence restricting the single use of nonce or even rekeying does not prevent the attacks.
%
% From the nature of the linear cryptanalysis, the attack works in the known plaintext attacks. Moreover, if a fixed plaintext is encrypted multiple times, the correlation can be used to recover a 1-bit information of the plaintext. The attack principally has larger impact because it can be used to recover more unknown bits of the plaintext, provided an initial segment of the plaintext is known. 
%
  To identify this correlation, we make use of rotational symmetries in \MORUS
  using linear masks that are invariant by word-rotations of the
  state.  This motivates us to introduce single-word versions of
  \MORUS called \MiniMORUS, which simplifies the
  analysis. The attack has been implemented and verified on
  \MiniMORUS, where it yields a correlation of $2^{-16}$.

>>>>

  As our main result, we present a
  linear correlation in the keystream of full \MORUS, which can be used to distinguish its output
  from random and to recover some plaintext bits in the
  broadcast setting.  For \MORUS[1280], the correlation is $2^{-76}$,
  which can be exploited after around $2^{152}$ encryptions, less than
  what would be expected for a 256-bit secure cipher.  For
  \MORUS[640], the same attack results in a correlation of $2^{-73}$, which does not
  violate the security claims of the cipher.

% More precisely, we show that a linear correlation exists between some bits of the plaintext and ciphertext. Moreover the bias depends purely on the plaintext, and not on the secret key nor the nonce, hence restricting the single use of nonce or even rekeying does not prevent the attacks.
%
% From the nature of the linear cryptanalysis, the attack works in the known plaintext attacks. Moreover, if a fixed plaintext is encrypted multiple times, the correlation can be used to recover a 1-bit information of the plaintext. The attack principally has larger impact because it can be used to recover more unknown bits of the plaintext, provided an initial segment of the plaintext is known. 
%
  To identify this correlation, we make use of rotational invariants in \MORUS
  using linear masks that are invariant by word-rotations of the
  state.  This motivates us to introduce single-word versions of
  \MORUS called \MiniMORUS, which simplifies the
  analysis. The attack has been implemented and verified on
  \MiniMORUS, where it yields a correlation of $2^{-16}$.

### 2018-09-06 AC2018_camera_ready/morusAC_01_Introduction.tex
Camera-ready version

To obtain this result, we propose a simplified abstraction of \MORUS, called \MiniMORUS. \MiniMORUS takes advantage of certain rotational symmetries in \MORUS and simplifies the description and analysis of the attack. We then show how the attack can be extended from \MiniMORUS to the real \MORUS.
%We observe that the actual bias of the approximation is better than suggested by the construction of the characteristics.
To confirm the validity of our analysis,
we practically verified the correlation of the full linear trail for \MiniMORUS, as well as the correlation of trail fragments for the full \MORUS.
Our analysis is also backed by a symbolic evaluation of the full trail equation and its correlation on all variants of \MORUS.

@@

\subsubsection*{Outline.}
This paper is organized as follows.
We first provide a brief description of \MORUS in \Cref{sec/Preliminaries}.
In \Cref{sec/introminimorus}, we introduce \MiniMORUS, an abstraction of \MORUS based on a certain class of rotational symmetries.
We analyze this simplified scheme in \Cref{sec/minitrails} and provide a ciphertext-only linear approximation with a weight of 16.
We then extend our result to the full scheme in \Cref{sec/fulltrails}, showing a correlation in the keystream over 5 steps,
and discuss the implications of our observation for the security of \MORUS in \Cref{sec/Discussion}.
In \Cref{sec/IniFin}, we present our results on the security of \MORUS with round-reduced initialization (in a nonce-misuse setting) or finalization.
We conclude in \Cref{sec/Conclusion}.

>>>>

To obtain this result, we propose a simplified abstraction of \MORUS, called \MiniMORUS. \MiniMORUS takes advantage of certain rotational invariants in \MORUS and simplifies the description and analysis of the attack. We then show how the attack can be extended from \MiniMORUS to the real \MORUS.
%We observe that the actual bias of the approximation is better than suggested by the construction of the characteristics.
To confirm the validity of our analysis,
we practically verified the correlation of the full linear trail for \MiniMORUS, as well as the correlation of trail fragments for the full \MORUS.
Our analysis is also backed by a symbolic evaluation of the full trail equation and its correlation on all variants of \MORUS.

@@

\subsubsection*{Outline.}
This paper is organized as follows.
We first provide a brief description of \MORUS in \Cref{sec/Preliminaries}.
In \Cref{sec/introminimorus}, we introduce \MiniMORUS, an abstraction of \MORUS based on a certain class of rotational invariants.
We analyze this simplified scheme in \Cref{sec/minitrails} and provide a ciphertext-only linear approximation with a weight of 16.
We then extend our result to the full scheme in \Cref{sec/fulltrails}, showing a correlation in the keystream over 5 steps,
and discuss the implications of our observation for the security of \MORUS in \Cref{sec/Discussion}.
In \Cref{sec/IniFin}, we present our results on the security of \MORUS with round-reduced initialization (in a nonce-misuse setting) or finalization.
We conclude in \Cref{sec/Conclusion}.

### 2018-09-06 AC2018_camera_ready/morusAC_02_Preliminaries.tex
Camera-ready version

\StateUpdate{} is composed of 5 rounds with similar operations.
The additional input $m^t$ is used in rounds 2 to 5, but not in round 1.
Each round uses the word-wise rotation (left circular shift) operation,
denoted $\lll_w$ in the following and \texttt{Rotl\_xxx\_yy} in the design document.
It divides a $q$-bit register value into 4 words of $w = q/4$ bits, and performs a rotation on each $w$-bit word.
The word-wise rotation constants $b_i$ for round $i$ are defined in \Cref{Tbl:rcon}.
Additionally, each round uses rotations on a whole $q$-bit register by a multiple of the word size,
denoted $\lll$ in the following and \texttt{<<<} in the design document.
The register-wise rotation constants $b_i'$ are also listed in \Cref{Tbl:rcon}.
%\todo{rounds $0\dots4$ or $1\dots5$?}

@@

\begin{table}[t]%[!bht]
  \centering
  \caption{Rotation constants $b_i$ for $\lll_w$ and $b_i'$ for $\lll$ in round $i$ of \texorpdfstring{\MORUS}{MORUS}.} \label{Tbl:rcon}
  \setlength{\tabcolsep}{6pt}
  \begin{tabular}{@{}l@{\qquad}rrrrr rrrrr@{}}
    \toprule
    & \multicolumn{5}{@{}c@{~}}{Word-wise rotation $\lll_w$} & \multicolumn{5}{@{}c@{}}{Register-wise rotation $\lll$} \\
    \cmidrule(r){2-6}
    \cmidrule(l){7-11}
    & $b_0$  & $b_1$  & $b_2$  & $b_3$  & $b_4$  
    & $b_0'$ & $b_1'$ & $b_2'$ & $b_3'$ & $b_4'$  \\
    %& \makebox[2em]{$b_0$} & \makebox[2em]{$b_1$} & \makebox[2em]{$b_2$} & \makebox[2em]{$b_3$} & \makebox[2em]{$b_4$} & \makebox[2em]{$b'_0$} & \makebox[2em]{$b'_1$} & \makebox[2em]{$b'_2$} & \makebox[2em]{$b'_3$} & \makebox[2em]{$b'_4$} \\ \hline
    \midrule
    \MORUS[640]  &  5 & 31 &  7 & 22 & 13 & 32 &  64 &  96 &  64 & 32 \\
    \MORUS[1280] & 13 & 46 & 38 &  7 &  4 & 64 & 128 & 192 & 128 & 64 \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\subsubsection{Finalization.}
The finalization phase generates the authentication tag $T$ using 10 more \StateUpdate{} steps. We only discuss the case where $T$ is not truncated. The associated data length and the message length are used to update the state:
\begin{enumerate}
\item $L \leftarrow |A| \mathrel\| |M|$ for \MORUS[640] or $L \leftarrow |A| \mathrel\| |M| \mathrel\|0^{128}$ for \MORUS[1280], where $|A|, |M|$ are represented as 64-bit integers.
\item $S^{u+v}_4 \leftarrow S^{u+v}_4 \oplus S^{u+v}_0.$
\item For $t = u+v, u+v+1, \ldots, u+v+9,$ compute
%\begin{equation*}
$S^{t+1} \leftarrow \StateUpdate{} (S^t, {\it tmp})$.
%\end{equation*}
%\item For \MORUS[640], $T = S^{u+v+10}_0 \oplus (S^{u+v+10}_1 \lll 96) \oplus ( S^{u+v+10}_2 \cdot S^{u+v+10}_3)$. For \MORUS[1280], $S^{u+v+10}_0 \oplus (S^{u+v+10}_1 \lll 192) \oplus ( S^{u+v+10}_2 \cdot S^{u+v+10}_3)$ is computed and the least significant 128 bits are produced as a tag $T$.
\item $T = S^{u+v+10}_0 \oplus (S^{u+v+10}_1 \lll b_2') \oplus ( S^{u+v+10}_2 \cdot S^{u+v+10}_3)$, or the least significant 128 bits of this value in case of \MORUS[1280].
\end{enumerate}

@@

%Finally, let $x$ denote a register. Recall that registers of \MORUS are composed of four words. We now define register-wise and word-wise rotations (denoted by \texttt{<<<} and \texttt{Rotl\_xxx\_yy} respectively in the \MORUS design document):\\
%\centerline{
%\def\arraystretch{1.25}
%\begin{tabular}{lcl}
%$x \rotl y$ &:& circular shift of $x$ by $y$ bits to the left.\\
%$x \rotlxy y$ &:& circular shift of each word within $x$ by $y$ bits to the left.\\
%\end{tabular}
%}
%
%The $b_i$'s denote rotation constants, whose value depends on which version of \MORUS is considered. Throughout, $w$ denotes the  word size, i.e. $w=32$ for \MORUS[640] and $w=64$ for \MORUS[1280].

>>>>

\StateUpdate{} is composed of 5 rounds with similar operations.
The additional input $m^t$ is used in rounds 2 to 5, but not in round 1.
Each round uses the bit-wise rotation (left circular shift) operation inside word,
denoted $\lll_w$ in the following and \texttt{Rotl\_xxx\_yy} in the design document.
It divides a $q$-bit register value into 4 words of $w = q/4$ bits, and performs a rotation on each $w$-bit word.
The bit-wise rotation constants $b_i$ for round $i$ are defined in \Cref{Tbl:rcon}.
Additionally, each round uses rotations on a whole $q$-bit register by a multiple of the word size,
denoted $\lll$ in the following and \texttt{<<<} in the design document.
The word-wise rotation constants $b_i'$ are also listed in \Cref{Tbl:rcon}.
%\todo{rounds $0\dots4$ or $1\dots5$?}

@@

\begin{table}[t]%[!bht]
  \centering
  \caption{Rotation constants $b_i$ for $\lll_w$ and $b_i'$ for $\lll$ in round $i$ of \texorpdfstring{\MORUS}{MORUS}.} \label{Tbl:rcon}
  \setlength{\tabcolsep}{6pt}
  \begin{tabular}{@{}l@{\qquad}rrrrr rrrrr@{}}
    \toprule
    & \multicolumn{5}{@{}c@{~}}{Bit-wise rotation $\lll_w$} & \multicolumn{5}{@{}c@{}}{Word-wise rotation $\lll$} \\
    \cmidrule(r){2-6}
    \cmidrule(l){7-11}
    & $b_0$  & $b_1$  & $b_2$  & $b_3$  & $b_4$  
    & $b_0'$ & $b_1'$ & $b_2'$ & $b_3'$ & $b_4'$  \\
    %& \makebox[2em]{$b_0$} & \makebox[2em]{$b_1$} & \makebox[2em]{$b_2$} & \makebox[2em]{$b_3$} & \makebox[2em]{$b_4$} & \makebox[2em]{$b'_0$} & \makebox[2em]{$b'_1$} & \makebox[2em]{$b'_2$} & \makebox[2em]{$b'_3$} & \makebox[2em]{$b'_4$} \\ \hline
    \midrule
    \MORUS[640]  &  5 & 31 &  7 & 22 & 13 & 32 &  64 &  96 &  64 & 32 \\
    \MORUS[1280] & 13 & 46 & 38 &  7 &  4 & 64 & 128 & 192 & 128 & 64 \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\subsubsection{Finalization.}
The finalization phase generates the authentication tag $T$ using 10 more \StateUpdate{} steps. We only discuss the case where $T$ is not truncated. The associated data length and the message length are used to update the state:
\begin{enumerate}
\item $L \leftarrow |A| \mathrel\| |M|$ for \MORUS[640] or $L \leftarrow |A| \mathrel\| |M| \mathrel\|0^{128}$ for \MORUS[1280], where $|A|, |M|$ are represented as 64-bit integers.
\item $S^{u+v}_4 \leftarrow S^{u+v}_4 \oplus S^{u+v}_0.$
\item For $t = u+v, u+v+1, \ldots, u+v+9,$ compute
%\begin{equation*}
$S^{t+1} \leftarrow \StateUpdate{} (S^t, L)$.
%\end{equation*}
%\item For \MORUS[640], $T = S^{u+v+10}_0 \oplus (S^{u+v+10}_1 \lll 96) \oplus ( S^{u+v+10}_2 \cdot S^{u+v+10}_3)$. For \MORUS[1280], $S^{u+v+10}_0 \oplus (S^{u+v+10}_1 \lll 192) \oplus ( S^{u+v+10}_2 \cdot S^{u+v+10}_3)$ is computed and the least significant 128 bits are produced as a tag $T$.
\item $T = S^{u+v+10}_0 \oplus (S^{u+v+10}_1 \lll b_2') \oplus ( S^{u+v+10}_2 \cdot S^{u+v+10}_3)$, or the least significant 128 bits of this value in case of \MORUS[1280].
\end{enumerate}

@@

%Finally, let $x$ denote a register. Recall that registers of \MORUS are composed of four words. We now define word-wise and bit-wise rotations (denoted by \texttt{<<<} and \texttt{Rotl\_xxx\_yy} respectively in the \MORUS design document):\\
%\centerline{
%\def\arraystretch{1.25}
%\begin{tabular}{lcl}
%$x \rotl y$ &:& circular shift of $x$ by $y$ bits to the left.\\
%$x \rotlxy y$ &:& circular shift of each word within $x$ by $y$ bits to the left.\\
%\end{tabular}
%}
%
%The $b_i$'s denote rotation constants, whose value depends on which version of \MORUS is considered. Throughout, $w$ denotes the  word size, i.e. $w=32$ for \MORUS[640] and $w=64$ for \MORUS[1280].

### 2018-09-06 AC2018_camera_ready/morusAC_03_MiniMorus.tex
Camera-ready version

\section{Symmetric Linear Combinations and \MiniMORUS}
\label{sec/introminimorus}

@@

\subsection{Symmetric Linear Combinations}

We begin with a few observations about the \StateUpdate{} function. Besides XOR and AND operations, the \StateUpdate{} function uses two types of bit rotations:
\begin{enumerate}
\item \emph{word-wise} rotations perform a circular shift on each word within a register;
\item \emph{register-wise} rotations perform a circular shift on a whole register.
\end{enumerate}
The second type of rotation always shifts registers by a multiple of the word size $w$. This amounts to a (circular) permutation of the words within the register: for example, if a register contains the words $(A, B, C, D)$, and a register-wise rotation by $w$ bits to the left is performed, then the register now contains the words $(B, C, D, A)$.

To build our linear trail, we start with a linear combinations of bits within a single register.
\begin{definition}
Recall that $w$ denotes the word size in bits, and $4w$ is the size of a register. A linear combination of the form:
\[
S^t_{i,j(0)} \oplus S^t_{i,j(1)} \oplus \dots \oplus S^t_{i,j(k)}
\]
is said to be \emph{symmetric} iff the set of bits $S^t_{i,j(0)}, \dots, S^t_{i,j(k)}$ is left invariant by a circular shift by $w$ bits; that is, iff:
\[
\{j(i) : i\leq k\} = \{j(i) + w \text{\rm{} mod } 4w : i\leq k\}.
\]
\end{definition}
\emph{Example.} The following linear combination is symmetric for \MORUS[640], i.e. $w = 32$:
\begin{equation}
S^t_{0,0} \oplus S^t_{0,32} \oplus S^t_{0,64} \oplus S^t_{0,96}.
\label{eq:symmetric}
\end{equation}

This definition naturally extends to a linear combination across multiple registers, and also across ciphertext blocks.
The value of such a linear combination is unaffected by register-wise rotations, since those rotations always shift registers by a multiple of the word size.
On the other hand, since word-wise rotations always shift all four words within a register by the same amount, word-wise rotations preserve the symmetry property. Moreover, the XOR of two symmetric linear combinations is also symmetric.%; and the same holds for the AND operation (if we extend the symmetric property to non-linear combinations in the natural way).

This naturally leads to the idea of building a linear trail using only symmetric linear combinations, which is what we are going to do. As a result, the effect of register-wise rotations can be ignored. Moreover, since all linear combinations we consider are going to be symmetric, they can be described by truncating the linear combination to the first word of a register. Indeed, an equivalent way of saying a linear combination is symmetric, is that it involves the same bits in each word within a register. For example, in the case of (\ref{eq:symmetric}) above, the four bits involved are the first bit of each of the four words.

@@

In fact, we can go further and consider a reduced version of \MORUS where each register contains a single word instead of four. The \StateUpdate{} function is unchanged, except for the fact that register-wise rotations are removed: see \Cref{fig:minimorus}. We call these reduced versions \MiniMORUS[640] and \MiniMORUS[1280], for \MORUS[640] and \MORUS[1280] respectively. Since registers in \MiniMORUS contain a single word, word-wise and register-wise rotations are the same operation; for simplicity we write $\lll$ for word-wise rotations.

Since the trail we are building is relatively complex, we will first describe it on \MiniMORUS. We will then extend it to the full \MORUS via the previous symmetry.

>>>>

\section{Rotational Invariant Linear Combinations and \MiniMORUS}
\label{sec/introminimorus}

@@

\subsection{Rotational Invariant Linear Combinations}

We begin with a few observations about the \StateUpdate{} function. Besides XOR and AND operations, the \StateUpdate{} function uses two types of bit rotations:
\begin{enumerate}
\item \emph{bit-wise} rotations perform a circular shift on each word within a register;
\item \emph{word-wise} rotations perform a circular shift on a whole register.
\end{enumerate}
The second type of rotation always shifts registers by a multiple of the word size $w$. This amounts to a (circular) permutation of the words within the register: for example, if a register contains the words $(A, B, C, D)$, and a word-wise rotation by $w$ bits to the left is performed, then the register now contains the words $(B, C, D, A)$.

To build our linear trail, we start with a linear combinations of bits within a single register.
\begin{definition}
Recall that $w$ denotes the word size in bits, and $4w$ is the size of a register. A linear combination of the form:
\[
S^t_{i,j(0)} \oplus S^t_{i,j(1)} \oplus \dots \oplus S^t_{i,j(k)}
\]
is said to be \emph{rotational invariant} iff the set of bits $S^t_{i,j(0)}, \dots, S^t_{i,j(k)}$ is left invariant by a circular shift by $w$ bits; that is, iff:
\[
\{j(i) : i\leq k\} = \{j(i) + w \text{\rm{} mod } 4w : i\leq k\}.
\]
\end{definition}
\emph{Example.} The following linear combination is rotational invariant for \MORUS[640], i.e. $w = 32$:
\begin{equation}
S^t_{0,0} \oplus S^t_{0,32} \oplus S^t_{0,64} \oplus S^t_{0,96}.
\label{eq:symmetric}
\end{equation}

This definition naturally extends to a linear combination across multiple registers, and also across ciphertext blocks.
The value of such a linear combination is unaffected by word-wise rotations, since those rotations always shift registers by a multiple of the word size.
On the other hand, since bit-wise rotations always shift all four words within a register by the same amount, bit-wise rotations preserve the rotational invariant property. Moreover, the XOR of two rotational invariant linear combinations is also rotational invariant.%; and the same holds for the AND operation (if we extend the symmetric property to non-linear combinations in the natural way).

This naturally leads to the idea of building a linear trail using only rotational invariant linear combinations, which is what we are going to do. As a result, the effect of word-wise rotations can be ignored. Moreover, since all linear combinations we consider are going to be rotational invariant, they can be described by truncating the linear combination to the first word of a register. Indeed, an equivalent way of saying a linear combination is rotational invariant, is that it involves the same bits in each word within a register. For example, in the case of (\ref{eq:symmetric}) above, the four bits involved are the first bit of each of the four words.

@@

In fact, we can go further and consider a reduced version of \MORUS where each register contains a single word instead of four. The \StateUpdate{} function is unchanged, except for the fact that word-wise rotations are removed: see \Cref{fig:minimorus}. We call these reduced versions \MiniMORUS[640] and \MiniMORUS[1280], for \MORUS[640] and \MORUS[1280] respectively. Since registers in \MiniMORUS contain a single word, bit-wise and word-wise rotations are the same operation; for simplicity we write $\lll$ for bit-wise rotations.

Since the trail we are building is relatively complex, we will first describe it on \MiniMORUS. We will then extend it to the full \MORUS via the previous rotational invariant property.
